Extended Data

### 2021-11-13 doc/manuscript/Dow_manuscript_Nature.Rmd
coauthor feedback

*NOTE: We're probably pretty close to the upper length limit. The typical length of an article with 3-4 modest display items (figures and tables) is 2000-2500 words (summary paragraph plus body text). We're currently at ~2512 words (1609 intro + results, discussion 903) and 1 modest + 2 large display items. Articles typically have 30-50 references (currently ok there)*

@@

In recent decades, Earth's forests have sequestered ~20% of anthropogenic CO~2~ emissions, thereby slowing the pace of atmospheric CO~2~ accumulation and climate change[@pan_large_2011].
A large portion of this CO~2~ sink occurs in temperate deciduous forests, with uptake of >100 Tg C yr^-1^ (>10% of the total forest C sink) attributable to environmental (as opposed to demographic) changes[@pugh_role_2019] *Coauthors suggest removing this but i'm unsure. I find this distinction relevant, but if more experienced people disagree I'm fine removing - would like your opinion*. 
The future behavior of this CO~2~ sink will play an important yet uncertain role in influencing atmospheric CO~2~ and climate change[@arora_carbon_2020; @friedlingstein_global_2020].

@@

While the responses of leaf phenology and seasonal *ecosystem (?)* CO~2~ sequestration to warming spring temperatures have been relatively well-studied[@jeong_phenology_2011; @keenan_net_2014; @buermann_widespread_2018; @ibanez_forecasting_2010; @menzel_european_2006; @menzel_growing_1999], we know very little about how the longest-lived component of fixed carbon in trees, the woody growth, is responding to warming spring temperatures. 
In fact, we know little about the stem-growth phenology of temperate deciduous species, with multi-year records of growth phenology having been published for only one site *(to our knowledge)*[@dorangeville_peak_2021]. <!-- I'm not aware of any others, but we do want to make sure we're not missing anything. klesse_integrating_2016 and martin_hydrometeorology_2017 both do this, but for conifers.-->
The climate sensitivity of woody growth phenology and its link to annual growth has never been studied *in-situ* (but see Ref.[@zohner_how_2021] for a controlled sapling experiment).

@@

Here, we evaluate how early spring temperatures affect stem growth phenology, growth rates, and annual growth of temperate deciduous trees in two forests and annual growth across forests spanning eastern North America. To test the assumption that warmer springs lead to an extension of the growing season we used measurements from 465 dendrometer bands across two decidious mid-atlantic forests. To test whether spring temperatrues were consistently positively correlated with annual ring-width Index (RWI; tree-ring width which has been standardized for age) we used `r n_chronologies` tree-ring chronologies from `r n_cores_sites` forests across eastern North America.

@@

Both MODIS-derived canopy foliage phenology and dendrometerband measurements of stem growth phenology generally shifted backwards as spring temperatures increased (Fig. `r DOY_timing`, `r EDF_pheno_all`). We found a consistent effect of temperature ($T_{max}$ or $T_{min}$) throughout the Spring, but the strongest effects on stem-growth phenology were found using $T_{max}$ during a critical temperatrue window (CTW). CTW was identified by measuring the correlation between all combinations of weekly $T_{max}$ and $DOY_{25}$ from January 1 and $DOY_{25}$ (`r EDF_climwin`). The CTW was defined as the week(s) which had the strongest correlation with $DOY_{25}$.

@@

To understand how annual growth increments have responded to spring temperatures at the centennial scale, we analyzed tree-ring chronologies of `r n_chronologies_SCBI` species at SCBI[@helcoski_growing_2019] and `r n_chronologies_HF` species at Harvard Forest (`r EDT_species`), along with an additional `r n_chronologies-n_chronologies_SCBI-n_chronologies_HF` chronologies from `r n_cores_sites-2` sites (Fig. `r tree_rings`; `r EDF_map`; `r EDT_tree_ring_summary`)[@dorangeville_drought_2018].
In total, our analysis included `r n_chronologies` chronologies representing `r n_cores_species` broadleaf species at `r n_cores_sites` sites distributed from Alabama (Lat = 34.35) to Michigan (Lat = 45.56) and spanning a 15 $^\circ$C range in April $T_{max}$. We converted the ring-width measurements into a dimensionless growth index (hereafter simply referred to as RWI; see Methods).
Across all chronologies, RWI was significantly (at p=0.05) positively correlated with April $T_{max}$ for only 1% of chronologies: 1 of 142 ring-porous and 2 of 66 diffuse-porous species-site combinations (`r EDT_tree_ring_summary`).
In contrast, RWI was frequently negatively correlated with $T_{max}$ during peak growing season months (May-August), with significant correlations for 53% (May: 47/142, Jun: 108/142, Jul: 90/142, Aug: 57/142) and 46% (May: 11/66, Jun: 53/66, Jul: 37/66, Aug: 22/66) of species-site-month combinations for ring-porous and diffuse-porous species, respectively.
$T_{min}$ generally exibited weaker relationships to annual growth than $T_{max}$, with few significant correlations between spring $T_{min}$ and RWI (`r EDF_tree_rings_TMIN`).

To test whether the negative effect of summer temperatures might offset the enhancement of growth by warmer spring temperatrues, we tested for the joint effects of April and June-July $T_{max}$ on RWI.
Results were qualitatively similar to the univariate correlations (Fig. `r tree_rings`), with significant (at p = 0.05) positive correlations to April $T_{max}$ for only 7% of chronologies and significant negative correlations with June-July $T_{max}$ for 77% of chronologies, supporting that summer temperatrues were the more important driver of annual stem growth (`r EDT_tree_ring_summary`).

@@

Together, our results demonstrate that warmer spring temperatures in the temperate deciduous forests of eastern North America advance the phenology of tree stem growth but have little effect on annual woody productivity (Figs. `r schematic_summary`- `r tree_rings`). 
The observed phenological advance in the start of stem growth under warmer springs parallels phenological advances observed for canopy foliage (Fig. `r DOY_timing`)[@keenan_net_2014;@finzi_carbon_2020;@stuble_plant_2021] and NEE [@keenan_net_2014; @finzi_carbon_2020].
However, inconsistent with the concept that an earlier start to growth would increase annual woody productivity, we demonstrate that warmer springs advance the end to the period of most active stem growth and have a negligible effect on total annual growth for most species and locations (Fig. `r tree_rings`). 
Our observations suggest that the cessation of rapid stem expansion, which occurs mid-summer near the time of peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020], is likely driven by cues other than carbon limitation, such as daylength or sink limitation, which also play an important role in autumn leaf senescence[@zani_increased_2020;@keenan_timing_2015;@fu_variation_2014].
Rather than lengthening the growing season, the primary effect of warming temperature on annual growth is a negative correlation with $T_{max}$ and associated drought stress during the peak growing season[@dorangeville_drought_2018], offsetting any benefit of an earlier onset of growth in the spring[@zhang_drought_2021].
Warm springs may also augment summer drought stress in some times and places, effectively canceling out any positive effects of an extended growing period[@buermann_widespread_2018]; however, spring temperatures and summer Standardized Precipitation Evapotranspiration Index[@vicente-serrano_multiscalar_2010] were uncorrelated within our dendrometer band analysis.

>>>>

*NOTE: We're probably pretty close to the upper length limit. The typical length of an article with 3-4 modest display items (figures and tables) is 2000-2500 words (summary paragraph plus body text). We're currently at ~2678 words (1609 intro + results, discussion 1069) and 1 modest + 2 large display items. Articles typically have 30-50 references (currently ok there)*

@@

In recent decades, Earth's forests have sequestered ~20% of anthropogenic CO~2~ emissions, thereby slowing the pace of atmospheric CO~2~ accumulation and climate change[@pan_large_2011].
A large portion of this CO~2~ sink occurs in temperate deciduous forests, with uptake of >100 Tg C yr^-1^ (>10% of the total forest C sink) attributable to environmental (as opposed to demographic) changes[@pugh_role_2019] <!-- Coauthors suggest removing this but i'm unsure. I find this distinction relevant, but if more experienced people disagree I'm fine removing - would like your opinion*! -->. 
The future behavior of this CO~2~ sink will play an important yet uncertain role in influencing atmospheric CO~2~ and climate change[@arora_carbon_2020; @friedlingstein_global_2020].

@@

While the responses of leaf phenology and seasonal *ecosystem (?)* CO~2~ sequestration to warming spring temperatures have been relatively well-studied[@jeong_phenology_2011; @keenan_net_2014; @buermann_widespread_2018; @ibanez_forecasting_2010; @menzel_european_2006; @menzel_growing_1999], we know very little about how the longest-lived component of fixed carbon in trees, the woody growth, is responding to warming spring temperatures. 
In fact, we know little about the stem-growth phenology of temperate deciduous species, with multi-year records of growth phenology having been published for only one site *(to our knowledge)*[@dorangeville_peak_2021]. <!-- I'm not aware of any others, but we do want to make sure we're not missing anything. klesse_integrating_2016 and martin_hydrometeorology_2017 both do this, but for conifers.-->
The climate sensitivity of woody growth phenology in deciduous trees and its link to annual growth has never been studied *in-situ* (but see Ref.[@zohner_how_2021] for a controlled sapling experiment).

@@

Here, we evaluate how early spring temperatures affect stem growth phenology, growth rates, and annual growth of temperate deciduous trees in two forests and annual growth across forests spanning eastern North America. To test the assumption that warmer springs lead to an extension of the growing season we used measurements from 465 dendrometer bands across two deciduous mid-atlantic forests. To test whether spring temperatures were consistently positively correlated with annual ring-width Index (RWI; tree-ring width which has been standardized for age) we used `r n_chronologies` tree-ring chronologies from `r n_cores_sites` forests across eastern North America.

@@

Both MODIS-derived canopy foliage phenology and dendrometer band measurements of stem growth phenology generally shifted backwards as spring temperatures increased (Fig. `r DOY_timing`, `r EDF_pheno_all`). We found a consistent effect of temperature ($T_{max}$ or $T_{min}$) throughout the Spring, but the strongest effects on stem-growth phenology were found using $T_{max}$ during a critical temperatrue window (CTW). CTW was identified by measuring the correlation between all combinations of weekly $T_{max}$ and $DOY_{25}$ from January 1 and $DOY_{25}$ (`r EDF_climwin`). The CTW was defined as the week(s) which had the strongest correlation with $DOY_{25}$.

@@

To understand how annual growth increments have responded to spring temperatures at the centennial scale, we analyzed tree-ring chronologies of `r n_chronologies_SCBI` species at SCBI[@helcoski_growing_2019] and `r n_chronologies_HF` species at Harvard Forest (`r EDT_species`), along with an additional `r n_chronologies-n_chronologies_SCBI-n_chronologies_HF` chronologies from `r n_cores_sites-2` sites (Fig. `r tree_rings`; `r EDF_map`; `r EDT_tree_ring_summary`)[@dorangeville_drought_2018].
In total, our analysis included `r n_chronologies` chronologies representing `r n_cores_species` broadleaf species at `r n_cores_sites` sites distributed from Alabama (Lat = 34.35) to Michigan (Lat = 45.56) and spanning a 15 $^\circ$C range in April $T_{max}$.
Across all chronologies, RWI was significantly (at p=0.05) positively correlated with April $T_{max}$ for only 1% of chronologies: 1 of 142 ring-porous and 2 of 66 diffuse-porous species-site combinations (`r EDT_tree_ring_summary`).
In contrast, RWI was frequently negatively correlated with $T_{max}$ during peak growing season months (May-August), with significant correlations for 53% (May: 47/142, Jun: 108/142, Jul: 90/142, Aug: 57/142) and 46% (May: 11/66, Jun: 53/66, Jul: 37/66, Aug: 22/66) of species-site-month combinations for ring-porous and diffuse-porous species, respectively.
$T_{min}$ generally exibited weaker relationships to annual growth than $T_{max}$, with few significant correlations between spring $T_{min}$ and RWI (`r EDF_tree_rings_TMIN`).

To test whether the negative effect of summer temperatures might offset the enhancement of growth by warmer spring temperatrues, we tested for the joint effects of April and June-July $T_{max}$ on RWI.
Results were qualitatively similar to the univariate correlations (Fig. `r tree_rings`), with significant (at p = 0.05) positive correlations to April $T_{max}$ for only 7% of chronologies and significant negative correlations with June-July $T_{max}$ for 77% of chronologies, supporting that summer temperatures were the more important driver of annual stem growth (`r EDT_tree_ring_summary`).

@@

Together, our results demonstrate that warmer spring temperatures in the temperate deciduous forests of eastern North America advance the phenology of tree stem growth but have little effect on annual woody productivity (Figs. `r schematic_summary`- `r tree_rings`). 
The observed phenological advance in the start of stem growth under warmer springs parallels phenological advances observed for canopy foliage (Fig. `r DOY_timing`)[@keenan_net_2014;@finzi_carbon_2020;@stuble_plant_2021] and NEE [@keenan_net_2014; @finzi_carbon_2020].
However, inconsistent with the concept that an earlier start to growth would increase annual woody productivity, we demonstrate that warmer springs advance the end to the period of most active stem growth and have a negligible effect on total annual growth for most species and locations (Fig. `r tree_rings`). 
Our observations suggest that the cessation of rapid stem expansion, which occurs mid-summer near the time of peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020], is likely driven by cues other than carbon limitation, such as daylength or sink limitation, which also play an important role in autumn leaf senescence[@zani_increased_2020;@keenan_timing_2015;@fu_variation_2014].
Rather than lengthening the growing season, the primary effect of warming temperature on annual growth is a negative correlation with $T_{max}$ and associated drought stress during the peak growing season[@dorangeville_drought_2018], offsetting any benefit of an earlier onset of growth in the spring[@zhang_drought_2021].
Warm springs may also augment summer drought stress in some times and places, effectively canceling out any positive effects of an extended growing period[@buermann_widespread_2018]; however, spring temperatures and summer Standardized Precipitation Evapotranspiration Index[@vicente-serrano_multiscalar_2010] were uncorrelated within our dendrometer band analysis.
The effect of warm springs specifically in years with higher levels of precipitation during the peak growing season is not elucidated from this study, and should be examined in the future to clarify this relationship.
